Cognitive Principles and Individual Differences \\ in Human Syllogistic Reasoning

### 2019-11-11 paper/paper-v3.tex
Noch kleinere Korrekturen/Anmerkungen

A conclusion then consists of an assertion about two properties over terms where each of these properties occurs in exactly one of the premises, and there is a third property connecting both premises which does not occur in the conclusion

@@

Clustering by Principles under WCS first presented in~\cite{declare:2017},
assumes that reasoning clusters can be expressed through cognitive principles and heuristic strategies. Cognitive principles are assumptions (not necessarily valid under CL) made by humans while reasoning, 
whereas heuristic strategies describe the methods that are applied when
conclusions are not derived by reasoning but instead based on structural patterns or {biases}.
This approach is an extension of the computational logic approach for human syllogistic reasoning presented in~\cite{cogsci:2017}: Three reasoning clusters and two heuristic strategies were specified, and achieved a match of 92\;\%, which was the highest result compared to the presented approaches in~\cite{khemlani:2012}.
% The system developed in~\cite{declare:2017}, was extended and participated at the Syllogism Challenge 2017.\footnote{\url{https://www.cc.uni-freiburg.de/modelingchallenge/challenge-2017}} 
Figure~\ref{overview:system:syllchal} illustrates the components of the system:
First, groups of reasoners were specified based on a set of underlying cognitive principles and heuristic strategies.
After that, this set was translated into a logic program (Section~\ref{sect:basic}-\ref{sect:advanced}), representing the group of reasoners.
Next, the respective models were computed (Section~\ref{sect:wcs}-\ref{sect:lfp}).
After that, the entailed relations between $a$ and $c$ were extracted (Section~\ref{sect:entail}). 
 The final answer distribution was specified according to the number of reasoning clusters 
that entailed the conclusion.
Finally, an illustrative representation of these clusters with respect to their distribution was done by means of multinomial process trees~\cite{Riefer1988}.

@@

Consider the following semantic operator~\cite{stenning:vanlambalgen:2008}: Let $I = \langle I^\top,I^\bot
\rangle$ be an interpretation. $\Phi_\CalP (I) =
\langle J^\top,J^\bot\rangle$, where
\[
\begin{array}{lcl}
J^\top & = & \{ A \mid A \leftarrow \Body \in \Def(A,\CalP)
\mbox{ and
} \Body \mbox{ is \True under } \langle I^\top,I^\bot \rangle \} \\
J^\bot & = & \{ A \mid \Def(A,\CalP) \neq \emptyset \mbox{ and }\\
&& \rule{7.5mm}{0mm}\Body \mbox{ is \False under } \langle
I^\top,I^\bot \rangle \mbox{ for all }  A \leftarrow \Body \in \Def(A,\CalP)\}.
\end{array}\Shrink
\]
The weak completion of non-contextual programs\footnote{Non-contextual programs refer programs that do not contain rules in which $\cxt$ appears in the body of any rule, i.e.\ $p = 0$.} 
always has a least model under  {\L}ukasiewicz logic,
which can be computed by the least fixed point of~$\Svlp$\cite{hk:2009a}. The correspondence to other well-known logic programming approaches, in particular 
to the well-founded semantics has been shown in~\cite{dietz:hoelldobler:wernhard:2014}.

The weak completion of contextual
programs might have more than one minimal model, and it is not guaranteed that 
the least fixed point of $\Svlp$ always exists. However,
for programs without cycles, there always exists a unique supported model,
which corresponds to the least fixed point of~$\Svlp$~\cite{dietz:hoelldobler:pereira:2017,dietz:hoelldobler:philipp:2017}. This model is \textit{supported} in the sense that 
it will always be computed by $\Svlp$, independent with which interpretation, we start to iterate $\Svlp$ with. Let us denote this model as $\CalM_\CalP$:
% As the fixed point of $\Svlp$ is unique and only computes the supported minimal model of $\WComp\CalP$, 
% we define 
$\CalP \ModelsWCS F$ iff $\CalM_\CalP(F) = \top$.
For all programs considered in the sequel, $\CalM_\CalP$ exists.

@@

\subsubsection{Converse Implication (\converse) and \\ Unknown Generalization (\unknownGen)}
Consider the statement \textit{Some y are z.} ({\MI yz}), which in CL, is represented as $\exists X (y(X) \wedge z(X))$
and which is semantically equivalent to $\exists X(z(X) \wedge y(X))$ (\converse).
As~\cite{khemlani:2012}
shows, humans seem to
distinguish between \textit{some y are z} and \textit{all y are z} (\unknownGen).
Accordingly, if we only observe that an object $o$ belongs to $y$ and $z$ then we do not want to conclude both, \textit{some y are z} and \textit{all y are z}.
In order to distinguish between \textit{some y are z} and \textit{all y are z}, we introduce the following principle:  If we know that \textit{some y are z}, then there must not only be an object~$o_1$ which belongs to $y$ and $z$, but there must be another object $o_2$ which belongs to
$y$ and which does not belong to~$z$.
$\CalP_{Iyz}$, the logic program representing (\MI yz), is as follows:\footnote{For simplicity, here we omit the clauses introduced by ({\footnotesize\textsf{converse}}). They can be found in~\cite{declare:2017}.}
\begin{align}
z(X) & \leftarrow y(X) \wedge \neg \Ab_{yz}(X). & \tag{\conditionals\ \&\ \licenses}\\
\Ab_{yz}(o_1) & \leftarrow  \bot. & \tag{\unknownGen\ \& \ \licenses}\\
y(o_1) & \leftarrow  \top. & \tag{\import} \\
y(o_2) & \leftarrow  \top. & \tag{\unknownGen} 
\end{align}
Consider the second clause: Different to the universal moods, \ME\ and \MA, $\Ab_{yz}$ is only false for $o_1$, which in turn implies that 
$\Ab_{yz}$ is unknown for all other objects in the program.
We obtain $\CalM_{\CalP_{Iyz}}= \langle \{y(o_1), y(o_2), z(o_1)\}, \{\Ab_{yz}(o_1)\} \rangle$.
One should observe that  $\Ab_{yz}(o_2)$ is an unknown assumption in
$\CalP_{Iyz}$ and,
hence, $\CalM_{\CalP_{Iyz}}(z(o_2)) = \udf$.

@@

We specify when $\MP$ entails a conclusion with respect to $yz$:\footnote{$yz$ can be replaced by $ac$ or~$ca$.}
\begin{description}
\item[\MA yz (all)] $\CalP \models Ayz$ iff there exists an object $o$ such that $\CalP \ModelsWCS
y(o)$ and for all objects $o$ we find that if $\CalP \ModelsWCS y(o)$ then $\CalP \ModelsWCS z(o)$.
\item [\ME yz (no)]
$\CalP \models Eyz$ iff there exists an object $o_1$ such that $\CalP \ModelsWCS
y(o_1)$ and for all objects $o_1$ we find that if $\CalP \ModelsWCS y(o_1)$ then
$\CalP \ModelsWCS \neg z(o_1)$.\footnote{
When the contraposition principles applies, then additionally 
the following condition needs to hold as well: There exists an object $o_2$ such that $\CalP \ModelsWCS
z(o_2)$ and for all objects $o_2$ we find that if $\CalP \ModelsWCS z(o_2)$ then
$\CalP \ModelsWCS \neg y(o_2)$.}
\item[\MI yz (some)]
$\CalP \models Iyz$ iff there exists an object $o_1$ such that $\CalP \ModelsWCS
y(o_1) \wedge z(o_1)$ and there exists an
object $o_2$ such that $\CalP \ModelsWCS y(o_2)$ and
$\CalP \not\ModelsWCS z(o_2)$ and
% there exists an object $o_3$ such that $\CalP \ModelsWCS
% z(o_3) \wedge y(o_3)$ and 
there exists an
object $o_3$ such that $\CalP \ModelsWCS z(o_3)$ and
$\CalP \not\ModelsWCS y(o_3)$.
\item[\MO yz (some are not)]
$\CalP \models Oyz$ iff there exists an object $o_1$ such that $\CalP \ModelsWCS
y(o_1) \wedge \neg z(o_1)$ and there exists an
object $o_2$ such that $\CalP \ModelsWCS y(o_2)$ and
$\CalP \not\ModelsWCS \neg z(o_2)$.
\item[\NVC\ (no valid conclusion)] iff none of the previous conclusions can be derived.
\end{description}
For \MI yz (some) the conversion is explicitly specified, motivated by the (\converse) principle.
By requiring for \MI yz (\MO yz, resp) the existence of an object with the first property $y$,
for which the second property $z$ has to be either either false or unknown (true or unknown, resp.),
\MA yz and \MI yz, (\ME yz and \MO yz, resp.), exclude each other.

@@

\cite{khemlani:2016} proposed an approach to model the individual reasoning patterns of the participants, and 
showed that different participants seem to derive different conclusions.
Their approach is built on top of the mReasoner~\cite{khemlani:2013}, an extension of the mental model theory~\cite{johnsonlaird:1983}.
Reasoning depends on the construction and manipulation of mental models, based on 3 fundamental principles: 
(i) represent possibilities, (ii) principle of iconicity, and (iii) principle of dual processes.
 The procedure of mReasoner consists of two steps:
(1) The construction of a small set of models that represents the terms of the premises, and
(2) the call of a deliberative component to search for counterexamples to conclusions.
They specified four parameters for the construction of a mental model and evaluated their
predictions based on 2 datasets, week 1 and week 2, for 20 participants.
The parameters for the construction of the mental models are (i) the 
size of a mental model, (ii) the probability of drawing a conclusions only from the most common mental models, or from the complete set,
(iii) the probability to search for counter examples,
and (iv) the probability to weaken the conclusion.
The clustering with mReasoner was based on a particular clustering algorithm to determine the optimal number of clusters for their dataset.
The K-means clustering algorithm was used to divide participants into clusters and then
the mReaoner to simulated the clusters by choosing appropriate parameter settings.
\cite{khemlani:2016} identified three different reasoning clusters: The intuitive cluster, the intermediate cluster and the deliberative cluster.
The approach is novel in the sense that until now no other cognitive theory tried to rigorously account for individual differences in human syllogistic reasoning. However, their approach seems to have two drawbacks. First, some of their 
parameters were of probabilistic nature, i.e.\ there is no explaining factor behind them. Second, the authors did not distinguish between the participants' 
responses in week 1 and week 2, but instead merged the responses. This might give a distorted picture, as participants could
have learned from week 1 to week 2.

@@

We have investigated the cognitive adequacy of clustering by principles under WCS with respect to the individual answer patterns reported by~\cite{khemlani:2016}.
\cite{khemlani:2016} is the only approach we are aware of that has evaluated predictions on human syllogistic reasoning
with respect to the non-aggregated data.
However in contrast to here, they did not distinguish between the participants' responses in week 1 and week 2.

>>>>

A conclusion then consists of an assertion about two properties over terms where each of these properties occurs in exactly one of the premises, and there is a third property connecting both premises which does not occur in the conclusion.

@@

Clustering by principles under WCS was first presented in~\cite{declare:2017};
the authors conjectured that reasoning clusters can be expressed through cognitive principles and heuristic strategies. Cognitive principles are assumptions (not necessarily valid under CL) made by humans while reasoning, 
whereas heuristic strategies describe the methods that are applied when
conclusions are not derived by reasoning but instead based on structural patterns or {biases}.
This approach is an extension of the computational logic approach for human syllogistic reasoning presented in~\cite{cogsci:2017}: Three reasoning clusters and two heuristic strategies were specified, and achieved a match of 92\;\%, which was the highest result compared to the presented approaches in~\cite{khemlani:2012}.
% The system developed in~\cite{declare:2017}, was extended and participated at the Syllogism Challenge 2017.\footnote{\url{https://www.cc.uni-freiburg.de/modelingchallenge/challenge-2017}} 
Figure~\ref{overview:system:syllchal} illustrates the components of the system:
First, groups of reasoners were specified based on a set of underlying cognitive principles and heuristic strategies.
After that, this set was translated into a logic program (Section~\ref{sect:basic}-\ref{sect:advanced}), representing the group of reasoners.
Next, the respective models were computed (Section~\ref{sect:wcs}-\ref{sect:lfp}).
After that, the entailed relations between $a$ and $c$ were extracted (Section~\ref{sect:entail}). 
 The final answer distribution was specified according to the number of reasoning clusters 
that entailed the conclusion.
Finally, an illustrative representation of these clusters with respect to their distribution was done by means of multinomial process trees~\cite{Riefer1988}.

@@

Consider the following semantic operator~\cite{stenning:vanlambalgen:2008}: Let $I = \langle I^\top,I^\bot
\rangle$ be an interpretation. $\Phi_\CalP (I) =
\langle J^\top,J^\bot\rangle$, where
\[
\begin{array}{lcl}
J^\top & = & \{ A \mid A \leftarrow \Body \in \Def(A,\CalP)
\mbox{ and
} \Body \mbox{ is \True under } \langle I^\top,I^\bot \rangle \} \\
J^\bot & = & \{ A \mid \Def(A,\CalP) \neq \emptyset \mbox{ and }\\
&& \rule{7.5mm}{0mm}\Body \mbox{ is \False under } \langle
I^\top,I^\bot \rangle \mbox{ for all }  A \leftarrow \Body \in \Def(A,\CalP)\}.
\end{array}\Shrink
\]
The weak completion of non-contextual programs\footnote{Non-contextual programs refer programs that do not contain rules in which $\cxt$ appears in the body of any rule, i.e.\ $p = 0$.} 
always has a least model under  {\L}ukasiewicz logic,
which can be computed by the least fixed point of~$\Svlp$~\cite{hk:2009a}. The correspondence to other well-known logic programming approaches, in particular 
to the well-founded semantics has been shown in~\cite{dietz:hoelldobler:wernhard:2014}.

The weak completion of contextual
programs might have more than one minimal model, and it is not guaranteed that 
the least fixed point of $\Svlp$ always exists. However,
for programs without cycles, there always exists a unique supported model,
which corresponds to the least fixed point of~$\Svlp$~\cite{dietz:hoelldobler:pereira:2017,dietz:hoelldobler:philipp:2017}. This model is \textit{supported} in the sense that 
it will always be computed by $\Svlp$, independent from the interpretation we start to iterate $\Svlp$ with. Let us denote this model as $\CalM_\CalP$:
% As the fixed point of $\Svlp$ is unique and only computes the supported minimal model of $\WComp\CalP$, 
% we define 
$\CalP \ModelsWCS F$ iff $\CalM_\CalP(F) = \top$.
For all programs considered in the sequel, $\CalM_\CalP$ exists.

@@

\subsubsection{Converse Implication (\converse) and \\ Unknown Generalization (\unknownGen)}
Consider the statement \textit{Some y are z.} ({\MI yz}), which in CL, is represented as $\exists X (y(X) \wedge z(X))$
and which is semantically equivalent to $\exists X(z(X) \wedge y(X))$ (\converse).
As shown by~\cite{khemlani:2012}, humans seem to
distinguish between \textit{some y are z} and \textit{all y are z} (\unknownGen).
Accordingly, if we only observe that an object $o$ belongs to $y$ and $z$ then we do not want to conclude both, \textit{some y are z} and \textit{all y are z}.
In order to distinguish between \textit{some y are z} and \textit{all y are z}, we introduce the following principle:  If we know that \textit{some y are z}, then there must not only be an object~$o_1$ which belongs to $y$ and $z$, but there must be another object $o_2$ which belongs to
$y$ and which does not belong to~$z$.
$\CalP_{Iyz}$, the logic program representing (\MI yz), is as follows:\footnote{For simplicity, here we omit the clauses introduced by ({\footnotesize\textsf{converse}}). They can be found in~\cite{declare:2017}.}
\begin{align}
z(X) & \leftarrow y(X) \wedge \neg \Ab_{yz}(X). & \tag{\conditionals\ \&\ \licenses}\\
\Ab_{yz}(o_1) & \leftarrow  \bot. & \tag{\unknownGen\ \& \ \licenses}\\
y(o_1) & \leftarrow  \top. & \tag{\import} \\
y(o_2) & \leftarrow  \top. & \tag{\unknownGen} 
\end{align}
Consider the second clause: Different to the universal moods, \ME\ and \MA, $\Ab_{yz}$ is only false for $o_1$, which in turn implies that 
$\Ab_{yz}$ is unknown for all other objects in the program.
We obtain $\CalM_{\CalP_{Iyz}}= \langle \{y(o_1), y(o_2), z(o_1)\}, \{\Ab_{yz}(o_1)\} \rangle$.
One should observe that  $\Ab_{yz}(o_2)$ is an unknown assumption in
$\CalP_{Iyz}$ and,
hence, $\CalM_{\CalP_{Iyz}}(z(o_2)) = \udf$.

@@

We specify when $\MP$ entails a conclusion with respect to $yz$:\footnote{$yz$ can be replaced by $ac$ or~$ca$.}
\begin{description}
\item[\MA yz (all)] $\CalP \models Ayz$ iff there exists an object $o$ such that $\CalP \ModelsWCS
y(o)$ and for all objects $o$ we find that if $\CalP \ModelsWCS y(o)$ then $\CalP \ModelsWCS z(o)$.
\item [\ME yz (no)]
$\CalP \models Eyz$ iff there exists an object $o_1$ such that $\CalP \ModelsWCS
y(o_1)$ and for all objects $o_1$ we find that if $\CalP \ModelsWCS y(o_1)$ then
$\CalP \ModelsWCS \neg z(o_1)$.\footnote{
When the contraposition principles applies, then additionally 
the following condition needs to hold as well: There exists an object $o_2$ such that $\CalP \ModelsWCS
z(o_2)$ and for all objects $o_2$ we find that if $\CalP \ModelsWCS z(o_2)$ then
$\CalP \ModelsWCS \neg y(o_2)$.}
\item[\MI yz (some)]
$\CalP \models Iyz$ iff there exists an object $o_1$ such that $\CalP \ModelsWCS
y(o_1) \wedge z(o_1)$ and there exists an
object $o_2$ such that $\CalP \ModelsWCS y(o_2)$ and
$\CalP \not\ModelsWCS z(o_2)$ and
% there exists an object $o_3$ such that $\CalP \ModelsWCS
% z(o_3) \wedge y(o_3)$ and 
there exists an
object $o_3$ such that $\CalP \ModelsWCS z(o_3)$ and
$\CalP \not\ModelsWCS y(o_3)$.
\item[\MO yz (some are not)]
$\CalP \models Oyz$ iff there exists an object $o_1$ such that $\CalP \ModelsWCS
y(o_1) \wedge \neg z(o_1)$ and there exists an
object $o_2$ such that $\CalP \ModelsWCS y(o_2)$ and
$\CalP \not\ModelsWCS \neg z(o_2)$.
\item[\NVC\ (no valid conclusion)] iff none of the previous conclusions can be derived.
\end{description}
For \MI yz (some) the conversion is explicitly specified, motivated by the (\converse) principle.
By requiring for \MI yz (\MO yz, resp.) the existence of an object with the first property $y$,
for which the second property $z$ has to be either either false or unknown (true or unknown, resp.),
\MA yz and \MI yz, (\ME yz and \MO yz, resp.), exclude each other.

@@

\cite{khemlani:2016}\todo{R: cite als Konstituent} proposed an approach to model the individual reasoning patterns of the participants, and 
showed that different participants seem to derive different conclusions.
Their approach is built on top of the mReasoner~\cite{khemlani:2013}, an extension of the mental model theory~\cite{johnsonlaird:1983}.
Reasoning depends on the construction and manipulation of mental models, based on 3 fundamental principles: 
(i) represent possibilities, (ii) principle of iconicity, and (iii) principle of dual processes.
 The procedure of mReasoner consists of two steps:
(1) The construction of a small set of models that represents the terms of the premises, and
(2) the call of a deliberative component to search for counterexamples to conclusions.
They\todo{R: oder doch ``Khemlani''? ;)} specified four parameters for the construction of a mental model and evaluated their
predictions based on 2 datasets, week 1 and week 2, for 20 participants.
The parameters for the construction of the mental models are (i) the 
size of a mental model, (ii) the probability of drawing conclusions only from the most common mental models, or from the complete set,
(iii) the probability to search for counter examples,
and (iv) the probability to weaken the conclusion.
The clustering with mReasoner was based on a particular clustering algorithm to determine the optimal number of clusters for their dataset.
The K-means clustering algorithm was used to divide participants into clusters and then
the mReaoner to simulated the clusters by choosing appropriate parameter settings.
\cite{khemlani:2016}\todo{R: cite als Konstituent} identified three different reasoning clusters: The intuitive cluster, the intermediate cluster and the deliberative cluster.
The approach is novel in the sense that until now no other cognitive theory tried to rigorously account for individual differences in human syllogistic reasoning. However, their approach seems to have two drawbacks. First, some of their 
parameters were of probabilistic nature, i.e.\ there is no explaining factor behind them. Second, the authors did not distinguish between the participants' 
responses in week 1 and week 2, but instead merged the responses. This might give a distorted picture, as participants could
have learned from week 1 to week 2.

@@

We have investigated the cognitive adequacy of clustering by principles under WCS with respect to the individual answer patterns reported by~\cite{khemlani:2016}.
\cite{khemlani:2016}\todo{R: cite als Konstituent} is the only approach we are aware of that has evaluated predictions on human syllogistic reasoning
with respect to the non-aggregated data.
However in contrast to here, they did not distinguish between the participants' responses in week 1 and week 2.
